---
title: "R Primer - WA_01"
subtitle: "DSC_406.001"
author: "Jeanne"
date: today 
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: right
theme:
  light: simplex
  dark: cyborg
editor: visual
---

## 0 - Introduction

This is a brief primer for R programming. We will have several of these assignments to do in class. What you do not finish in class you will complete out of class by the due date. I encourage you to work together 😀!

This primer will use RStudio with Quarto.

"[Quarto](https://quarto.org/docs/get-started/hello/rstudio.html#:~:text=Quarto%20is%20a%20multi%2Dlanguage,to%20use%20RStudio%20with%20Quarto.) is a multi-language, next-generation version of R Markdown from Posit and includes dozens of new features and capabilities while at the same being able to render most existing Rmd files without modification."

### How to use this document

This is an Quarto Notebook.

A key way to use it is to be sure that you are viewing the document in the "Visual Editor" mode. You can use this mode by clicking the symbol that appears like a letter A (or the tip of a pencil!) in the top right of this window.

![](img/visual_1.svg)

Let's get started!

### 🤿 Let's dive in.......

### Entering commands in R

"Chunks", "block", or "blob" of R code can be added

-   using the Add Chunk command in the editor toolbar
-   the CTRL + ALT + I shortcut (OS X: Cmd + Option + I)
-   or by typing the chunk delimiters `{r} and`.)

#### **👉 Your Turn** **⤵**

Add a code chunk with a method of your choice.

```{r}

```

#### Code Chunk Options:

| Option  | Description                                                                                                                                                                                                                          |
|---------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| eval    | Evaluate the code chunk (if false, just echos the code into the output)                                                                                                                                                              |
| echo    | Include the source code in output output Include the results of executing the code in the output (true, false, or asis to indicate that the output is raw markdown and should not have any of Quarto's standard enclosing markdown). |
| warning | Include warnings in the output.                                                                                                                                                                                                      |
| error   | Include errors in the output (note that this implies that errors executing code will not halt processing of the document).                                                                                                           |
| include | Catch all for preventing any output (code or results) from being included (e.g. `include: false` suppresses all output from the code block).                                                                                         |

Example:

```{r}

#| eval:false
#| warning: false

str(cars)

```

### Execute the code:

-   Each line of R can be executed by clicking on the line and pressing CTRL + ENTER shortcut
-   Or, you can press the green triangle on the right-hand side to run everything in the chunk

Click the green arrow in the right corner of the "chunk"-of code to execute the "Hello World" script.

```{r}
#This is a comment
print("Hello World")
```

### **Objects**

A variable is a letter or word which takes (or contains) a value. We use the assignment 'operator', `<-` to create a variable and store some value in it.

```{r}
#This is a comment
#store 10 in x and then inspect x 
x <- 10
x
```

```{r}
myNumber <- 25
myNumber
```

We also can perform arithmetic on variables using functions:

```{r}
sqrt(myNumber)
```

We can add variables together:

```{r}
x + myNumber
```

We can change the value of an existing variable:

```{r}
x <- 21
x
```

We can set one variable to equal the value of another variable:

```{r}
x <- myNumber
x
```

When we are feeling lazy we might give our variables short names (x, y, i...etc), but a best practice would be to give them meaningful names. - Always lower case - Cannot start with a number or contain characters such as ., \_, '-'. - No spaces use underscore to connect - DO not name variables the same as as in-built functions in R, such as c, T, mean - Don't ever have names like TRUE/FALSE or NaN, inf, NA

#### **👉 Your Turn** **⤵**

First, Assign a number between 1 and 10 to a new object call my_number. Second, Multiply my_number by 2 and save as my_product. Third, Type my_product and run to examine the contents.

```{r}
#ADD YOUR CODE BELOW


```

### Functions

Functions in R perform operations on arguments (the inputs(s) to the function).

mean() max() filter() class()

#### **👉 Your Turn** **⤵**

In the code chunk below, use the round() function to round 3.14 to the nearest whole number. Run your code to see the result.

```{r}
#ADD YOUR CODE BELOW


```

### Arguments

Arguments are always contained in parentheses -- curved brackets, () -- separated by commas.

Arguments can be named or unnamed, but if they are unnamed they must be ordered (we will see later how to find the right order). The names of the arguments are determined by the author of the function and can be found in the help page for the function. When testing code, it is easier and safer to name the arguments.

Use the ? to read about the arguments you can add to the 'mean()' function

?mean() ?class()

Take a look at the arguments fo rthe `round()` function.

#### **👉 Your Turn** **⤵**

Add an argument to the round() function to round pi to the nearest 1 decimal places. Remember to add a comma between arguments. *#HINT: look under USAGE*

```{r}
#ADD YOUR CODE BELOW


```

### Vectors

A vector is simply a list of items that are of the same type. To combine the list of items to a vector, use the c() function and separate the items by a comma. In the example below, we create a vector variable called fruits.

```{r}
# Vector of strings
fruits <- c("banana", "apple", "orange")

# Inspect fruits
fruits
```

You can add a constant to each element of the vector

```{r}
x <- 1:5
2 + x
```

You can multiply a constant by each element of a vector

```{r}
x <- 1:5
2 * x
```

You can do arithmetic on two vectors of the same length

```{r}
x <- c(1, 3, 5)
y <- c(2, 4, 6)
x + y

```

Or multiply

```{r}
x <- c(1, 3, 5)
y <- c(2, 4, 6)
x * y
```

We can create a vector that contains the numbers 2 through 10.

```{r}
vector <- c(2:10)
vector
```

You can create a vector that has logical values

```{r}
# Vector of logical values
log_values <- c(TRUE, FALSE, TRUE, FALSE)

log_values
```

#### **👉 Your Turn** **⤵**

Create a vector that contains the numbers 50-100, Save as a new object called my_vector.

```{r}
#ADD YOUR CODE BELOW


```

### Indexing

There are multiple ways to access or replace values in vectors or other data structures. The most common approach is to use "indexing". This is also referred to as "slicing".

In the below, note that brackets \[ \] are used for indexing, whereas you have already seen that parentheses ( ) are used to call a function. Later on, you will also see the use of { }. It is very important not to mix these up.

Lets create a vector

```{r}
b <- 10:15

#call the object

b
```

Get the first element of a vector

```{r}
b[1]
```

#### **👉 Your Turn** **⤵**

Get the second elements of a vector

```{r}

```

Now we can get elements 2 & 3

```{r}
b[2:3]

```

We can also use the c() function

```{r}
b[c(2,3)]
```

More advance, return all elements except the second

```{r}
b[-2]
```

#### **👉 Your Turn** **⤵**

We can use the c() function to return all elements except the second

```{r}
# this is the same using a c( ) function

```

### Data Types

Basic data types in R can be divided into the following types:

-   `numeric` - (10.5, 55, 787)
-   `integer` - (1L, 55L, 100L, where the letter "L" declares this as an integer)
-   `complex` - (9 + 3i, where "i" is the imaginary part)
-   `character` - (a.k.a. string) - ("k", "R is exciting", "FALSE", "11.5")
-   `logical` - (a.k.a. boolean) - (TRUE or FALSE)
-   `factor`- (a.k.a. categorical data of levels), factor(c("yes", "no", "yes"))

We can use the `class()` function to check the data type of a variable.

#### **👉 Your Turn** **⤵**

-   check the class of `log_values`
-   check the class of `vector`

```{r}
#ADD YOUR CODE BELOW



```

### Packages 📦

R uses "packages," add-ons that enhance its functionality. One package that we'll be using is the tidyverse. The {tidyverse} package is actually a [collection of R](https://www.tidyverse.org/packages) packages designed for reading, wrangling, and exploring data and which all share an underlying design philosophy, grammar, and data structures.

Before we can begin using these packages, we will need to install them using the `install.packages()` function built into R. Since we only need to install once, we install a packagein the console.

Click the green arrow in the right corner of the block-or "chunk"-of code that follows and see if you can identify which packages have been installed in the console below.

We will only install.packages(" ") in the console since they only need to be added once.

```{r}

#| eval:false

install.packages("tidyverse")
```

Once these packages have been installed, we will need to load them in order to use the handy functions they contain.

To load the tidyverse, click the green arrow in the right corner of the block-or "chunk"-of code that follows. Notice that we do not need to use the quotation marks again because the {tidyverse} package and packages it contains are now a part of our package library!

You will need to call your packages for every script.

```{r}
library(tidyverse)
```

Please do not worry if you saw a number of messages: those mean that the tidyverse loaded just fine. If you see an error, though, try to interpret or search via your search engine the contents of the error, or reach out to us for assistance.
